\bf{Inequality in non-cognitive skills: \\ a cross-cohort comparison}

### 2018-05-31 codes/R/slides/cohorts_tapes.Rnw
slides update after Hoynes questions

<<READ, cache=FALSE >>=
library(here)
read_chunk(here("codes/R/1_load.R"))
read_chunk(here("codes/R/2_efa.R"))
read_chunk(here("codes/R/3_inv_main.R"))
read_chunk(here("codes/R/3_inv_aux.R"))
read_chunk(here("codes/R/3_inv_fit.R"))
read_chunk(here("codes/R/4_scores.R"))
read_chunk(here("codes/R/4_scores_aux.R"))
read_chunk(here("codes/R/5_graphs.R"))
read_chunk(here("codes/R/6_selectvars.R"))
read_chunk(here("codes/R/8_regs_dep.R"))
read_chunk(here("codes/R/8_regs_outc.R"))
read_chunk(here("codes/R/getpars.R"))

@@

\begin{frame}
\frametitle{Motivation}
\framesubtitle{Inequality in skills}
\bi
\item Socio-emotional skills developed during childhood are important for lifetime outcomes
\item Socioeconomic gradients in these skills are persistent 
\item {\color{red} IMPROVE INTRO...}
\ei
\end{frame}

@@

\bi
  \item Usually, socioemotional constructs are measured using validated scales
    \bi \item e.g. Rutter, BSAG, SDQ \ei
  \item Most economic literature uses simple sum scores of these scales
  \item Sometimes useful to consider different populations/groups
    \bi
    \item Different countries
    \item Different subgroups (gender, race, ...)
    \item Same population, at different points in time
    \ei
  \item Do these scales measure the same constructs?
  \item Are the constructs measured in the same way?
  \item And what if only info from different scales is available?
\ei
\end{frame}

\begin{frame}
\frametitle{This study}
\framesubtitle{Research questions}
We compare the evolution of socioemotional skills in the UK
\bi
\item \textbf{Q1}: \textbf{How many latent dimensions} do we need to capture socioemotional skills in the data?
~\\
\pause
\item \textbf{Q2}: Can we derive \textbf{comparable measures} of socioemotional skills across cohorts of children born in different years?
~\\
\pause
\item \textbf{Q3}: How has \textbf{inequality} in early-life socioemotional skills evolved in the UK between 1975 and 2005?
~\\
\pause
\item \textbf{Q4}: Have the \textbf{determinants} of socioemotional skills evolved?
~\\
\pause
\item \textbf{Q5}: How do early socioemotional skills map into \textbf{adolescent/adult outcomes}?
\ei
\end{frame}

@@

\bi
\item Focus on socioemotional skills in two British cohorts
  \bi \item British Cohort Study (BCS, b. 1970)
  \item Millennium Cohort Study (MCS, b. 2000/1) \ei
~\\ \pause
\item \textbf{Contribution}
\bi
\item Construct a common 11-item scale, starting from different survey instruments
\item Investigate its psychometric properties, and find
  \bnm 
  \item \textbf{\textit{Bidimensionality}} -- Externalising and Internalising
  \item \textbf{\textit{Comparability}} in terms of scale (but not location)
  \enm
\ei
~\\ \pause

\item \textbf{Key findings}
  \bi
  \item Inequality in non-cognitive skills has increased in 1975-2005
  \item Sharpest change for boys rather than girls
  \item {\color{red} OTHER RESULTS}
  \ei
\ei

@@

\bi
\item Four main cohort studies:
  \bi
  \item \textit{National Survey of Health and Development} (NSHD) -- b. March 1946
  \item \textit{National Child Development Study} (NCDS) -- b. March 1958
  \item \textit{British Cohort Study} (BCS) -- b. April 1970
  \item \textit{Millennium Cohort Study} (MCS) -- b. 2000-2001
  \ei
\item Nationally representative samples of 10,000+ individuals from birth
  \bi
  \item Perinatal period (pregnancy, birth outcomes)
  \item Skill formation -- cognitive, socioemotional
  \item Health and health behaviours
  \item Family background and composition
  \item Schooling and labour market outcomes
  \ei
\pause
\vspace{0.2cm}
\item Not always \textbf{consistent}
  \bi
  \item Different design, timing, questions, survey instruments...
  \item Harmonisation effort underway (CLOSER)
  \ei
\ei

@@

\bi
\item Rutter and SDQ are different instruments
  \bi \item Not ideal to just use sum scores and assume they return comparable information... \ei
\item Focus on subset of items that have similar wording between the two instruments
\item $\Rightarrow$ New 11-item subscale
\ei

@@

\begin{frame}
\frametitle{Factor structure}
\framesubtitle{Existing evidence}

\textbf{Q1}: How many dimensions underlie the new 11-item scale?
\bi
\item Rutter A scale (BCS) 
  \bi 
  \item Traditionally conceptualised as bidimensional \citep{Blanden2007}
    \bi
    \item \emph{anti-social} -- or aggressive-hyperactive
    \item \emph{neurotic} -- or anxious-fearful
    \ei
  \item Sometimes the former is split into aggressive/conduct vs hyperkinetic \citep{McGee1985,Collishaw2004}
  \ei
\item Strengths and difficulties (MCS) 
  \bi
  \item Five subscales (emotional, hyperactivity, prosocial, peer, conduct) \citep{Stone2010}
  \item Evidence that lower-dimensional structures can work \citep{Goodman2010}
  \ei
\ei

@@

\begin{frame}
\frametitle{EFA}\label{efa}
\framesubtitle{Dimensionality}

\bi
\item Different methods for assessing number of factors give varied results \hyperlink{numfac}{\beamerbutton{Det.}}
\item We choose two factors to maximise interpretability
\item In both cohorts, EFA delivers a clear-cut split between two factors
\item Factors can be coherently interpreted as externalising ($\sim$ anti-social) and internalising ($\sim$ neurotic)
\ei 
\vspace{12pt}
\centering
\begin{spacing}{1.1}
\begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
  & \textbf{{\color{OliveGreen}Externalising}}  &    & \textbf{{\color{Orange}Internalising}}  \\
1 & Restless                & 7  & Worried                \\
2 & Squirmy/fidgety         & 8  & Afraid of new things   \\
3 & Fights/bullies children & 9  & Solitary               \\
4 & Cannot settle           & 10 & Unhappy/tearful        \\
5 & Temper tantrums         & 11 & Head/stomachaches \\
6 & Disobedient             &    &                       
\end{tabular}
\end{spacing}

@@

\bi
\item Two constructs isolated in the new scale, separately by cohort
\item But are externalising and internalising measured in a comparable way \emph{across cohorts}?
\ei
\vspace{12pt}\pause
\textbf{Measurement Invariance (MI)} analysis \citep{Wu2016a}
  \bi
  \item Formal test of the invariance hypothesis using factor analysis
  \item Aim: assess if constructs are measured in the same way in 2+ groups
  \item Procedure:
    \bnm
    \item Start from the most general model -- parameters allowed differ across groups
    \item Impose restrictions on sets of parameters to be equal across groups
    \item Test relative fit of restricted models compared to initial model
    \enm
  \ei

@@

\begin{frame}
\frametitle{Measurement Invariance}\label{midet}
\framesubtitle{IFA model}
\setstretch{1.1}

Categorical Item Factor Analysis (IFA) model
\bi
\item Latent propensity to endorse each item $i$ for child $j$ in group $c$ \citep{Muthen1984}
\begin{align}\label{eq:propens}
\begin{split}
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
\end{split}
\end{align}

\item Mean/covariance structure
$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
with $$(\theta^{EXT}_{jc}, \theta^{INT}_{jc}) = \bm{\Theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$

\item Discreteness
$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.

\ei
\end{frame}

\begin{frame}
\frametitle{Measurement Invariance}\label{misteps}
\framesubtitle{Steps \hyperlink{miapp1}{\beamerbutton{Details}}}

We follow the approach in \cite{Wu2016a}
\bi
\item Define group ($c$) as cohort-gender cell
  \bi \item Invariance results do not vary if separating by gender \ei
\ei
\vspace{.2cm}
\bnm
\item \textbf{Configural} model
  \bi
  \item Minimally restrictive (all parameters free across groups)
  \item Normalised on latent means $\bm{\kappa}$ and variances $\bm{\Phi}$
  \ei
\item \textbf{Threshold and Loading invariance} model
  \bi
  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, and $\bm{\Lambda}$ to be equal across groups
  \item Allows comparison of latent factor variances
  \ei
\item \textbf{Threshold, Loading, and Intercept invariance} model
  \bi
  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, $\bm{\Lambda}$, and $\bm{\nu}$ to be equal across groups
  \item Allows comparison of latent factor variances \emph{and} means
  \ei
\enm
\end{frame}

@@

\bi
\item We use RMSE to compare fit between models of different invariance level
\item Restricting thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit details}}
  \bi \item We can compare the scale (variance) of the latent factors \ei
\item Models with restricted intercepts do \emph{not} pass MI test
  \bi \item We cannot compare the location of latent factors in different groups \ei
\ei
\begin{center}
\resizebox{\textwidth}{!}{
<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
require(xtable)

@@

print(xtable(table,
                    align="L{0cm}L{.35\\textwidth}C{.12\\textwidth}C{.15\\textwidth}C{.1\\textwidth}C{.15\\textwidth}C{.1\\textwidth}",
                    digits = 4
                    ),
             size="\\footnotesize",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
            " & & \\multicolumn{2}{c}{\\textbf{Full sample}} & \\multicolumn{2}{c}{\\textbf{59-61 months}} \\\\ \\cmidrule(lr){3-4}  \\cmidrule(lr){5-6}
            Model & Num. par. & $\\chi^2$ & RMSEA & $\\chi^2$ & RMSEA \\\\"
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@@

\begin{frame}
\frametitle{Measurement Invariance}
\framesubtitle{Factor scores}
\bi
\item Nevertheless, we can use the model with non-invariant intercepts to score the latent factors (externalising and internalising dimensions)
\item Estimated parameters: \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}
\item Use Empirical Bayes approach to predict individual scores
\item All scores are coded so that \emph{higher is better}
\item \textbf{CAUTION}: location differences by cohort are not interpretable!
~\\
\pause
\item Is this an improvement on using simple sum scores?
  \bi
  \item Sum scores assume that each item conveys the same information
  \item Scores obtained from the model exploit the covariance structure of the whole 11-item scale
  \item Every 11-dimensional pattern of response in the data can be assigned a different score
  \item Substantial increase in variability ({\color{red} could add boxplot of factor scores vs sum scores})
  \ei
\ei
\end{frame}

@@

\bi
\item \textbf{Q3}: How has socioeconomic inequality in socioemotional skills increased between 1975 (BCS) and 2005 (MCS)?
\item 
\ei

@@

print(xtable(table,
                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
                    digits = 3
                    ),
             size="\\scriptsize",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,4)),
                              command = c(
                              # header
            "Measure & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
                       & BCS & MCS & BCS & MCS  \\\\
                      \\textbf{Externalising} \\\\ ",
                      "\\textbf{Internalising} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@@

\begin{frame}
\frametitle{Inequality}\label{ineq_sm}
\framesubtitle{Smoking in pregnancy}
\vspace{-.2cm}
\begin{figure}
<<FACINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=11, out.width='\\linewidth'>>=
# plot
fi <- fi_sm # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\end{figure}
\end{frame}

@@

\bi
\item Unconditionally, the distribution is more spread out (except for girls' internalising)
\item Inequality across socioeconomic indicators has increased, especially
  \bi
  \item for boys rather than girls
  \item considering inequality across maternal schooling and employment, rather than paternal occupation
  \ei
\ei

@@

We explain externalising and internalising skills using information from the birth and 5 year surveys \\

\begin{table}
\footnotesize
\begin{tabular}{L{.25\textwidth}C{.1\textwidth}L{.5\textwidth}}
Group & Age &  \\
\midrule
Father occupation & 5 & White collar (SC I-IIINM) \newline Blue collar (SC IIIM-V-other) \newline No father figure \\
Maternal education & 5 & Post-compulsory schooling \\
Maternal employment & 5 & Employed \\
Maternal background & 0/5 & Age, height \newline Nonwhite ethnicity (child) \newline Number of children in HH \\
Pregnancy & 0 & Firstborn child, number of previous stillbirths \newline Mother smoked in pregnancy \newline Preterm birth, (log) birth weight \\
\end{tabular}
\end{table}

@@

@@

write up Gelbach approach...

@@

\begin{frame}
\frametitle{Decomposition (EXT)}\label{gelb_ext_f}
\framesubtitle{Explaining the gradient by maternal compulsory schooling \hyperlink{gelb_ext}{\beamerbutton{Table}}}

\begin{frame}
\frametitle{Decomposition (INT)}\label{gelb_int_f}
\framesubtitle{Explaining the gradient by maternal compulsory schooling \hyperlink{gelb_int}{\beamerbutton{Table}}}

@@

\section{Mapping skills to later outcomes}

<<LOAD_DATA_OUTC, dependson="SELECTVARS">>=
# rload outcomes data
@

@@

\begin{frame}
\frametitle{Mapping skills into behaviours}
\framesubtitle{Adolescence (MCS)}
\centering
\setstretch{1.0}
\resizebox{\textwidth}{!}{
<<REGS_OUTC_MCS, results='asis', dependson="REGRESSIONS">>=

# collapse to matrix
mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab,matrix,ncol=11,byrow=FALSE))
table <- mcs_outctab_mat

# NAs in columns and rows with COG
table[,c(6,11)] <- NA
table[seq(4,16,4),] <- NA

>>>>

<<READ, cache=FALSE >>=
library(here)
read_chunk(here("codes/R/1_load.R"))
read_chunk(here("codes/R/2_efa.R"))
read_chunk(here("codes/R/3_inv_main.R"))
read_chunk(here("codes/R/3_inv_aux.R"))
read_chunk(here("codes/R/3_inv_fit.R"))
read_chunk(here("codes/R/4_scores.R"))
read_chunk(here("codes/R/4_scores_aux.R"))
read_chunk(here("codes/R/5_graphs.R"))
read_chunk(here("codes/R/6_selectvars.R"))
read_chunk(here("codes/R/8_regs_dep.R"))
read_chunk(here("codes/R/8_regs_outc.R"))
read_chunk(here("codes/R/9_gelbfig.R"))
read_chunk(here("codes/R/getpars.R"))

@@

\begin{frame}
\frametitle{Motivation}
\framesubtitle{Inequality in socioemotional skills}
\bi
\item Socioemotional skills developed during childhood are important for lifetime outcomes \citep{Lundberg2017}
\item The UK has witnessed sharp increases in income inequality and decreases in social mobility since the 1970s
\item What happened to inequalities in early childhood development in this period?
\ei
\end{frame}

@@

\bi
  \item Accurate, consistent measurement of skills is fundamental
  \item Usually, socioemotional constructs are measured using validated scales
    \bi 
    \item e.g. Rutter, BSAG, SDQ 
    \item Obtained by aggregating different items
    \ei
  \item Can we use these scales to compare different cohorts, by gender?
  \item And what if only info from different scales is available?
    \bi
    \item Can different scales measure the \textbf{same constructs}?
    \item Can the constructs be measured \textbf{in comparable ways}?
    \ei
\ei
\end{frame}

\begin{frame}
\frametitle{This study}
\framesubtitle{Research questions}
We compare the evolution of inequality in socioemotional skills at age 5 across two UK cohorts born 30 years apart
\bi
\item \textbf{Q1}: \textbf{How many latent dimensions} underlie the available survey instruments?
~\\
\pause
\item \textbf{Q2}: Can we derive \textbf{comparable measures} of socioemotional skills?
~\\
\pause
\item \textbf{Q3}: How has \textbf{inequality} in early-life socioemotional skills evolved?
~\\
\pause
\item \textbf{Q4}: Have the \textbf{determinants} of socioemotional skills changed?
~\\
\pause
\item \textbf{Q5}: Has the mapping of socioemotional skills into \textbf{adolescent/adult outcomes} changed?
\ei
\end{frame}

@@

\bi
\item Focus on socioemotional skills in two British cohorts 
  \bi \item British Cohort Study (b. 1970)
  \item Millennium Cohort Study (b. 2000/1) \ei
\item \textbf{Methodological contribution}
  \bi
  \item Construct a common 11-item scale from different survey instruments
    \bnm 
    \item \textbf{\textit{Bidimensionality}} -- Externalising and Internalising
    \item \textbf{\textit{Comparability}} in terms of variance (but not mean)
    \enm
  \ei
\item \textbf{Key findings}
  \bi
  \item Inequality in non-cognitive skills has increased in 1975-2005
  \item Sharpest change for boys rather than girls
  \item Increased importance of pregnancy factors, decreased importance of paternal education, in explaining skill inequality by maternal education
  \item In both cohorts, socioemotional skills (especially externalising) are significant predictors of social and health outcomes in adolescence
  \ei
\ei
\end{frame}

@@

\bi
\item Four main cohort studies:
  \bi
  \item \textit{National Survey of Health and Development} (NSHD) -- b. March 1946
  \item \textit{National Child Development Study} (NCDS) -- b. March 1958
  \item \textit{British Cohort Study} (BCS) -- b. April 1970
  \item \textit{Millennium Cohort Study} (MCS) -- b. 2000-2001
  \ei
\item Nationally representative samples of 10,000+ individuals from birth
  \bi
  \item Perinatal period (pregnancy, birth outcomes)
  \item Skill formation -- cognitive, socioemotional \citep{Joshi2014}
  \item Health and health behaviours
  \item Family background and composition
  \item Schooling and labour market outcomes
  \ei
\pause
\vspace{0.2cm}
\item Not always \textbf{consistent}
  \bi
  \item Different design, timing, questions, survey instruments...
  \item Harmonisation effort underway (CLOSER)
  \ei
\ei

@@

\bi
\item Rutter and SDQ are different instruments
  \bi \item Not ideal to just use sum scores and assume they return comparable information... \ei
\item Focus on subset of items that have similar wording between the two instruments
\item Recode items to enhance comparability when needed
  \bi \item E.g. ``Generally obedient" dichotomised to match ``Often disobedient"  \ei
\item $\Rightarrow$ New 11-item subscale
\ei

@@

\begin{frame}
\frametitle{Exploratory FA}\label{efa1}
\framesubtitle{Dimensionality}

\textbf{Q1}: How many dimensions underlie the new 11-item scale?

\bi
\item Factor analysis routinely used to construct measures of socioemotional skills \citep{Conti2014,Attanasio2017c}
\item Starts with \emph{exploratory} step
\item Decompose the covariance matrix of items to assess:
  \bi \item the dimensionality of the underlying factor structure (how many latent factors)
  \item which items measure which latent factor \ei
\item Psychometric literature often finds \textbf{two latent factors} provide good synthesis for Rutter and SDQ \hyperlink{efalit}{\beamerbutton{Det.}}
\item This can be tested in our data
\ei

@@

\begin{frame}
\frametitle{Exploratory FA}\label{efa2}
\framesubtitle{Dimensionality}

\textbf{Q1}: How many dimensions underlie the new 11-item scale?

\bi
\item Dimensionality
  \bi
  \item Methods for assessing dimensionality imply 1-3 factors \hyperlink{numfac}{\beamerbutton{Det.}}
  \item We choose \textbf{two factors} to maximise interpretability
  \ei
\item Structure
  \bi
  \item Exploratory FA using weighted least squares and \textit{oblimin} rotation
  \item Factors can be coherently interpreted as \textbf{externalising} ($\sim$ anti-social) and \textbf{internalising} ($\sim$ neurotic)
  \ei
\ei 
\pause
\vspace{6pt}
\centering
\small
\begin{spacing}{1.0}
\begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
  & \textbf{{\color{OliveGreen}Externalising}}  &    & \textbf{{\color{Orange}Internalising}}  \\
1 & Restless                & 7  & Worried                \\
2 & Squirmy/fidgety         & 8  & Afraid of new things   \\
3 & Fights/bullies children & 9  & Solitary               \\
4 & Cannot settle           & 10 & Unhappy/tearful        \\
5 & Temper tantrums         & 11 & Head/stomachaches \\
6 & Disobedient             &    &                       
\end{tabular}
\end{spacing}

@@

\bi
\item Two constructs uncovered in the new scale, separately by cohort
\item But are externalising and internalising measured in a comparable way \emph{across cohorts}?
\ei
\vspace{12pt}\pause
\textbf{Measurement Invariance (MI)} analysis
  \bi
  \item Formal test of the invariance hypothesis using factor analysis
  \item Aim: assess if constructs are measured in the same way in 2+ groups
  \item Procedure:
    \bnm
    \item Start from the most general model -- parameters allowed to differ across groups
    \item Impose restrictions on sets of parameters to be equal across groups
    \item Test relative fit of restricted models compared to initial model
    \enm
  \ei

@@

\begin{frame}
\frametitle{Measurement Invariance}\label{midet}
\framesubtitle{IFA model \hyperlink{midet_app}{\beamerbutton{Details}}}
\setstretch{1.1}

Categorical Item Factor Analysis (IFA) model
\bi
\item Observed categorical items $X_1$-$X_{11}$ measuring latent factor $\theta$
\item Latent propensity to endorse each item $i$ for child $j$ in group $c$ \citep{Muthen1984}
$$ X^*_{ijc} =\nu_{ci} + \lambda_{ci}\theta_{jc} + u_{ijc} $$
\item Discreteness \citep{Piatek2016}
$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
\vspace{.2cm}
\item We follow the approach in \cite{Wu2016a}
  \bi
  \item Define group ($c$) as cohort-gender cell
  \ei
\ei
\end{frame}

\begin{frame}[t]
\frametitle{Measurement Invariance}\label{misteps}
\framesubtitle{Steps \hyperlink{miapp1}{\beamerbutton{Details}}}
$$ X^*_{ijc} =
\only<1>{\textcolor{red}{\nu_{ci}} + \textcolor{red}{\lambda_{ci}}\theta_{jc} + u_{ijc}} 
\only<2>{\textcolor{red}{\nu_{ci}} + \textcolor{green}{\lambda_{i}}\theta_{jc} + u_{ijc}} 
\only<3>{\textcolor{green}{\nu_{i}} + \textcolor{green}{\lambda_{i}}\theta_{jc} + u_{ijc}} 
$$
$$X_{ijc} = s \qquad \text{if} \; 
\only<1>{\textcolor{red}{\tau_{s,ic}} \leq X^*_{ijc} < \textcolor{red}{\tau_{s+1,ic}}}
\only<2>{\textcolor{green}{\tau_{s,i}} \leq X^*_{ijc} < \textcolor{green}{\tau_{s+1,i}}}
\only<3>{\textcolor{green}{\tau_{s,i}} \leq X^*_{ijc} < \textcolor{green}{\tau_{s+1,i}}}
\qquad \text{for} \; s=0,1,2$$
\vspace{-.2cm}
\bnm
\item \textbf{Configural} model
  \bi
  \item Minimally restrictive (all parameters free across groups)
  \ei
  \pause
\item \textbf{Threshold and Loading invariance} model
  \bi
  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, and $\bm{\Lambda}$ to be equal across groups
  \item Allows comparison of latent factor variances
  \ei
  \pause
\item \textbf{Threshold, Loading, and Intercept invariance} model
  \bi
  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, $\bm{\Lambda}$, and $\bm{\nu}$ to be equal across groups
  \item Allows comparison of latent factor variances \emph{and} means
  \ei
\enm
\end{frame}

@@

\bi
\item We use RMSE to compare fit between models with different invariance restriction
\item Restricting only thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit details}}
  \bi \item We can compare the scale (variance) of the latent factors \ei
\item Models with restricted intercepts do \emph{not} pass MI test
  \bi \item We cannot compare the location of latent factors in different groups \ei
\ei
\begin{center}
\resizebox{\textwidth}{!}{
<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
require(xtable)

@@

print(xtable(table,
                    align="L{0cm}L{.35\\textwidth}C{.12\\textwidth}C{.15\\textwidth}C{.1\\textwidth}C{.15\\textwidth}C{.1\\textwidth}",
                    digits = 4
                    ),
             size="\\footnotesize",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
            " & & \\multicolumn{2}{c}{\\textbf{Full sample}} & \\multicolumn{2}{c}{\\textbf{59-61 months} \\hyperlink{agehist}{\\beamerbutton{Hist.}}} \\\\ \\cmidrule(lr){3-4}  \\cmidrule(lr){5-6}
            Model & Num. par. & $\\chi^2$ & RMSEA & $\\chi^2$ & RMSEA \\\\"
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@@

\begin{frame}
\frametitle{Measurement Invariance}\label{miscores}
\framesubtitle{Factor scores}

\textbf{Q2}: Can we derive \textbf{comparable measures} of socioemotional skills?
\bi
\item We can use the model with non-invariant intercepts to score the latent factors (externalising and internalising dimensions)
\item Estimated parameters: \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}
\item Use Empirical Bayes approach to predict individual scores \citep{Wu2013}
\item All scores are coded so that \emph{\textbf{a higher value means better socioemotional skills}}
\item \textbf{CAUTION}: location differences by cohort/gender are not interpretable!
\ei
\end{frame}

\begin{frame}
\frametitle{Measurement Invariance}\label{miscores2}
\framesubtitle{Factor scores vs. Sum scores}
\bi
\item Is this an improvement on using simple sum scores?
  \bi
  \item Sum scores assume that each item conveys the same information, and that everything is invariant
  \item Scores obtained from the model exploit the covariance structure of the whole 11-item scale
  \item Every 11-dimensional pattern of response in the data can be assigned a different score
  \item Substantial increase in variability \hyperlink{facraw}{\beamerbutton{Graph}}
  \ei
\ei
\end{frame}

@@

\bi
\item Previous work in the mental health literature has mostly looked at secular trends in socioemotional skills
  \bi 
  \item Focus on adolescent period \citep{Collishaw2004,Collishaw2010a}
  \item Mixed findings -- often stable for boys, increasing for girls \citep{Maughan2005,Bor2014}
  \ei
\item Less attention devoted to secular changes in 
  \bi
  \item socioemotional skills in the pre-school period
  \item socioeconomic inequality in skills
  \item early-life determinants of skills
  \ei
\item We look at
  \bi 
  \item Distribution (quantile differences)
  \item Mean skills by parents' socioeconomic background
  \ei
\ei

@@

print(xtable(table,
                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
                    digits = 2
                    ),
             size="\\scriptsize",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,4)),
                              command = c(
                              # header
            "Measure & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
                       & BCS & MCS & BCS & MCS  \\\\
                      \\textbf{Externalising} \\\\ ",
                      "\\textbf{Internalising} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@@

\begin{frame}
\frametitle{Inequality}\label{ineq_sm}
\framesubtitle{Smoking in pregnancy}
\vspace{-.2cm}
\begin{figure}
<<FACINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
# plot
fi <- fi_sm # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\end{figure}
\end{frame}

@@

\bi
\item Unconditionally, the distribution is more spread out (except for girls' internalising)
\item Inequality across socioeconomic indicators has increased, especially
  \bi
  \item for boys rather than girls
  \item considering inequality by maternal schooling and employment, rather than paternal occupation
  \ei
\ei

@@

\begin{frame}
\frametitle{Determinants of socioemotional skills}\label{detintro}

\textbf{Q4}: Have the \textbf{determinants} of socioemotional skills changed across cohorts?
\bi 
\item Previous work has mostly focused on single cohorts, and not estabilished invariance
\item \cite{McCulloch2000,Blanden2008,Fitzsimons2017}
\item Linear regression (Tobit results are very similar)
\item Merge information from the birth and 5 year surveys \hyperlink{detvars}{\beamerbutton{Variables}}
\ei
\end{frame}

@@

\textbf{Q4}: Have the \textbf{determinants} of socioemotional skills changed across cohorts?

\bi
\item Diminishing importance of father's occupation in determining skills
\item Increased importance of
  \bi 
  \item Maternal education and employment
  \item Smoking in pregnancy
  \item Birthweight
  \ei
\ei

@@

\bi
\item We decompose the difference between children of mothers with compulsory education vs. post-compulsory
\item Use the same groups of covariates as above
\item Approach by \cite{Gelbach2016}
  \bi
  \item Based on simple omitted variable bias expression
  \item Invariant to the ordering of variables
  \ei
\ei

@@

<<GELB_FIG>>=
# make gelbach graphs
@

\begin{frame}
\frametitle{Decomposition (EXT)}\label{gelb_ext_f}
\framesubtitle{Maternal compulsory schooling \hyperlink{gelb_ext}{\beamerbutton{Table}}}
\begin{figure}
<<GELB_FIG_EXT, dependson="GELB_FIG", dev='tikz', fig.height=4.5, fig.width=9, out.width='\\linewidth'>>=
# plot
gelbplot_ext
@
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Decomposition (INT)}\label{gelb_int_f}
\framesubtitle{Maternal compulsory schooling \hyperlink{gelb_int}{\beamerbutton{Table}}}
\begin{figure}
<<GELB_FIG_INT, dependson="GELB_FIG", dev='tikz', fig.height=4.5, fig.width=9, out.width='\\linewidth'>>=
# plot
gelbplot_int
@
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Decomposition}
\bi
\item Differences in age 5 skills by mother's education have increased between 1975 and 2005
\item Observed variables can explain around half of the differentials
\item Father's occupation explains less in 2005 than 1975, while pregnancy factors explain more
\ei

@@

\section{Mapping skills to adolescent/adult outcomes}

\begin{frame}
\frametitle{Mapping skills to adolescent/adult outcomes}

\textbf{Q5}: Has the mapping of socioemotional skills into \textbf{adolescent/adult outcomes} changed?

\bi
\item We examine how sociemotional skills at age 5 map into outcomes later in the life course
  \bi 
  \item BCS: Age 16, 30, 38, 42
  \item MCS: Age 14 (last available survey)
  \ei
\item Previous work has mostly focused on single cohorts\cite{Carneiro2007,Conti2010a,Joshi2014}
\item Focus on behaviours and labour market outcomes
\pause
\item Include a \textbf{cognitive factor} at age 5, using available tests
  \bi 
  \item BCS: English Picture Vocabulary Test, Human Figure Drawing, Copy Designs
  \item MCS: British Ability Scales (Naming Vocabulary, Picture Similarities, Pattern Construction)
  \item Bartlett scoring
  \ei
\ei
\end{frame}

<<LOAD_DATA_OUTC, dependson="SELECTVARS">>=
# load outcomes data
@

@@

\begin{frame}
\frametitle{Mapping skills}
\framesubtitle{Adolescent behaviours (MCS)}
\begin{table}
\centering
\setstretch{1.0}
\resizebox{\textwidth}{!}{
<<REGS_OUTC_MCS, results='asis', dependson="REGRESSIONS">>=

# collapse to matrix
mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab[c("smktry14", "bmi14")],matrix,ncol=11,byrow=FALSE))
table <- mcs_outctab_mat
